Cleaning and Processing

### 2024-08-16 code/analysis-code/fullanalysis-v1.qmd
Revisions for submission 3
Clean up of code

## Merge Models 3 and 4
```{r}
# Combine the two tables using tbl_merge
table4 <- tbl_merge(list(table_model3, table_model4), tab_spanner = c("**Model 3**", "**Model 4**")) %>%
  modify_table_body(fun = ~.x %>% arrange(variable))
table4

>>>>

## Merge Models 3 and 4

```{r}
# Combine the two tables using tbl_merge
table4 <- tbl_merge(list(table_model3, table_model4), tab_spanner = c("**Model 3**", "**Model 4**")) %>%
  modify_table_body(fun = ~.x %>% arrange(variable))
table4

### 2024-08-16 code/analysis-code/introanalysis-v1.qmd
Revisions for submission 3
Clean up of code

---
title: "Intro Statistical Analysis Script"
author: "Cassia Roth"
date: "`r Sys.Date()`"
output: html_document
---

@@

```{r}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_linear.rds")
#load data
ML_linear <- readRDS(data_location)
```

## Linear Models

Let's run a simple regression with the outcome variable (birth weight in grams) and the exposure variable maternal ancestry. I will use the recipe and workflow features of `tidymodels`.

```{r}
# Model using recipes and workflows
# Define recipe for model
ML_recipe1 <- recipe(Weightgrams ~ ModifiedColor, data = ML_linear) %>% 
  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
  step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable

@@

# Summarize model fit with tidiers
ML_summary1 <- summary(ML_fit1) # Summary of fit
tidy(ML_fit1, conf.int = TRUE) # Coefficients and confidence intervals
glance(ML_fit1) # Model-level statistics

# Produce predictions for R-squared
ML_fit1_pred <- predict(ML_fit1, new_data = ML_linear %>% select(-Weightgrams))
ML_fit1_pred

@@

#Print table
table_fit1

# Save regression output into table
saveRDS(table_fit1, file = here::here("results", "tables", "table_fit1.rds"))
```
Now let's run a second simple bivariate analysis, looking at infant birth weight as a function of maternal age. This time, when plotting weight as a function of maternal age, we can use a scatterplot.

```{r}
# Model using recipes and workflows
# Create recipe
ML_recipe_age <- recipe(Weightgrams ~ Age, data = ML_linear)

@@

# Summarize model fit with tidiers
ML_summary_age <- summary(ML_fit_age) # Summary of fit
tidy(ML_fit_age, conf.int = TRUE) # Coefficients and confidence intervals
glance(ML_fit_age) # Model-level statistics

# Produce predictions for RMSE and R-squared
ML_age_pred <- predict(ML_fit_age, new_data = ML_linear %>% select(-Weightgrams))
ML_age_pred

@@

Now let's run a slightly more complicated model that includes maternal skin color, age, nationality, and gestational status.

@@

# Summarize model fit with tidiers
ML_lm3_summary <- summary(ML_lm3) # Summary of fit
tidy(ML_lm3, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
glance(ML_lm3) # Model-level statistics

@@

For all three models, the r-squared is very small (0.009, 0.008, and 0.027), indicating that the model does not explain much of the variance in birth weight. This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors.

@@

@@

```{r}
# Simple linear model
# Set reference level for Color
ML_linear$ModifiedNationality <- relevel(ML_linear$ModifiedNationality, ref = "Brazilian")

@@

# Summarize model fit with tidiers
ML_lm4_summary <- summary(ML_lm4) # Summary of fit
tidy(ML_lm4, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
glance(ML_lm4) # Model-level statistics

@@

# Save regression output into table
saveRDS(table_fit4, file = here::here("results", "tables", "table_fit4.rds"))
```
From our results, we can see that Europeans had heavier infants than Brazilians. Since all women of color in this sample were Brazilian, we have a problem with multicollinearity. So let's try a couple things to understand this issue. We will create a `Brazilian` (yes, no) variable and then run a regression with this variable. Then, we will drop all non-Brazilians from the study and run the original regression with the covariates.

@@

str(ML_linear)
```
Now, we will run a regression with only this new variable.

```{r}
# Simple linear model
# Set reference level for Brazilian
ML_linear$Brazilian <- relevel(ML_linear$Brazilian, ref = "Brazilian")

@@

# Summarize model fit with tidiers
ML_lm5_summary <- summary(ML_lm5) # Summary of fit
tidy(ML_lm5, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
glance(ML_lm5) # Model-level statistics

@@

# Save regression output into table
saveRDS(table_fit5, file = here::here("results", "tables", "table_fit5.rds"))
```
Here, we can see that non-Brazilians have babies on average weighing 100 grams more than Brazilians (p-value 0.001). So we will now run a regression of only Brazilian women using Afro-Descent and Euro-descent categories.

@@

# Simple linear model
# Set reference level for Brazilian
ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")

@@

# Summarize model fit with tidiers
ML_lm6_summary <- summary(ML_lm6) # Summary of fit
tidy(ML_lm6, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
glance(ML_lm6) # Model-level statistics

@@

Moving on from issues of collinearity, let's try to run a model where skin color is differentiated according to the original data, or with three categories (White, Black, and Mixed Race). We'll start with a simple model of infant birth weight as a function of maternal skin color. Then we will run a model with all the predictors.

```{r}
# Simple linear model
# Set reference level for Color
ML_linear$Color <- relevel(ML_linear$Color, ref = "White")

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm5 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color, data = ML_linear)

# Summarize model fit with tidiers
ML_lm5_summary <- summary(ML_lm5) # Summary of fit
tidy(ML_lm5, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
glance(ML_lm5) # Model-level statistics

# Create a gtsummary table directly from the model
table_fit5 <- tbl_regression(ML_lm4, exponentiate = FALSE, intercept = TRUE) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))
table_fit5

# Save regression output into table
saveRDS(table_fit5, file = here::here("results", "tables", "table_fit5.rds"))

@@

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm6 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear)

# Summarize model fit with tidiers
ML_lm6_summary <- summary(ML_lm6) # Summary of fit
tidy(ML_lm6, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
glance(ML_lm6) # Model-level statistics

@@

Now we will still consider `Birthweight` as the outcome of interest, but we will transform it into a categorical outcome of <=2500 (low birthweight, LBW) and >2500 (normal birthweight, NBW) and fit a logistic model, using the main predictor of interest, `ModifiedColor`.

First, we will split data into training and testing sets to later calculate ROC_AUC.

@@

Now we will run the first logistic model with `ModifiedColor` as the predictor.

```{r}
# Create recipe
# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
log1_recipe <- recipe(BirthweightCategory ~ ModifiedColor, data = ML_linear) #%>%
  #step_dummy(all_nominal(), -all_outcomes()) %>%
  #step_normalize(all_predictors())

@@

# Print the plot
print(roc_plot1)
```
With an ROC_AUC of 0.463, our model is performing worse than random guessing. Our pseudo R-squared is 0.004.

Now let's run another logistic model with `Color` as the predictor.

```{r}
# Simple logistic regression model with Color as the predictor
# Check factor levels
levels(ML_linear$BirthweightCategory) # Reference is LBW

@@

# Calculate metrics using pseudo-R2
# Extract model metrics using broom
model_metrics2 <- glance(log2_fitted)

@@

# Print the plot
print(roc_plot2)
```
The ROC_AUC for our second simple logistic regression model is 0.47, or still worse than random guessing. Our pseudo R-squared is 0.006.

Now, we will run two more logistic models, one with `ModifiedColor` and the other covariates `Age`, `ModifiedStatus`, `ModifiedNationality`, and one with `Color` and these same covariates.

```{r}
# Multiple logistic regression model with ModifiedColor as the predictor
# Check factor levels
levels(ML_linear$BirthweightCategory) # Reference is LBW
levels(ML_linear$ModifiedColor) # Reference is Euro-descent
levels(ML_linear$ModifiedStatus) # Reference is multiparous
levels(ML_linear$ModifiedNationality) # Reference is European

@@

# Print the plot
print(roc_plot3)
```
Our ROC_AUC here is worse, at 0.46, or worse than random guessing. Our pseudo-R2 is 0.0055.

```{r}
# Multiple logistic regression model with Color as the predictor
# Check factor levels
levels(ML_linear$BirthweightCategory) # Reference is LBW
levels(ML_linear$Color) # Reference is White
levels(ML_linear$ModifiedStatus) # Reference is multiparous
levels(ML_linear$ModifiedNationality) # Reference is European

@@

# Print the plot
print(roc_plot4)
```
In our final model, our ROC_AUC is 0.477, still worse than random guessing. Our pseudo-R2 is 0.007.

@@

# Save regression output into table
saveRDS(table5, file = here("results", "tables", "table5_final.rds"))
```
```

>>>>

---
title: "Intro Statistical Analysis Script"
author: "Cassia Roth"
date: "`r Sys.Date()`"
format: 
  html:
    code-fold: true
editor: visual
---

@@

```{r}
#Path to linear data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_linear.rds")
#load data
ML_linear <- readRDS(data_location)

#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_summary.rds")
#load data
ML_summary <- readRDS(data_location)
```

# LBW and NBW percentages

First, I will see how many infants were LBW or NBW for each maternal descendant category. Then, I will perform a simple difference in means (t-test) to see if the LBW and NBW means stratified by maternal skin color are statistically significant.

```{r}
# See LBW and NBW with true denominator
LBW_percentage <- ML_summary %>%
  filter(!is.na(ModifiedColor) & !is.na(BirthweightCategory)) %>% # Remove NAs from both variables

group_by(ModifiedColor) %>% # Group by ModifiedColor
# Calculate the percentage for each group
summarise(
  total_count = n(),
  lbw_count = sum(BirthweightCategory == "LBW"),
  lbw_percentage = (lbw_count / total_count) * 100
  )

# Print the result
print(LBW_percentage)

# Perform difference in means test (t-test) for birth weight category
# Modify code to create dataset for t-test
LBW_data <- ML_summary %>%
  filter(!is.na(ModifiedColor) & !is.na(BirthweightCategory)) %>%
  mutate(is_LBW = as.numeric(BirthweightCategory == "LBW"))  # 1 if LBW, 0 if not

# Print the first few rows to check the structure
print(head(LBW_data))

# Perform t-test
t_test_result <- t.test(is_LBW ~ ModifiedColor, data = LBW_data, subset = ModifiedColor %in% c("Euro-Descent", "Afro-Descent"))

# Print the result
print(t_test_result)

# Extract specific values for inline text
p_value <- t_test_result$p.value
t_statistic <- t_test_result$statistic
df <- t_test_result$parameter
mean_diff <- diff(t_test_result$estimate)
ci_lower <- t_test_result$conf.int[1]
ci_upper <- t_test_result$conf.int[2]
```

After performing a Welch Two Sample t-test to compare the rates of LBW between Euro- and Afro-descent mothers, the results were as follows: t = `r round(t_statistic, 2)`, df = `r round(df, 1)`, p-value = `r round(p_value, digits = 3)`, mean difference = `r round(mean_diff, 4)`, 95% CI = [`r round(ci_lower, 4)`, `r round(ci_upper, 4)`].

# Linear Models

To be able include metrics like $(R^2)$ dynamically in the Quarto document, I first need to store the values in a list. I will create a list to store the values. In each code chunk with a fitted model, I will add the $(R^2)$ to the list. Then I can use the stored values inline in the document.

```{r}
# Create a list to store R-squared values
r_squared_values <- list()
```

I will now run a simple regression with the outcome variable (birth weight in grams) and the exposure variable maternal ancestry. I will use the recipe and workflow features of `tidymodels`.

```{r}
# Model 1 using recipes and workflows
# Define recipe for model
ML_recipe1 <- recipe(Weightgrams ~ ModifiedColor, data = ML_linear) %>% 
  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level
  step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable

@@

# Summarize model fit with tidiers
ML_summary1 <- summary(ML_fit1) # Summary of fit
tidy(ML_fit1, conf.int = TRUE) # Coefficients and confidence intervals
model_stats1 <- glance(ML_fit1) # Extract model-level statistics

# This code is to write dynamic inline text in Quarto
# Extract R-squared value
r_squared_values$model1 <- model_stats1$r.squared

# This code is for performance measures that aren't necessary in this study
# Produce predictions for R-squared
ML_fit1_pred <- predict(ML_fit1, new_data = ML_linear %>% select(-Weightgrams))
ML_fit1_pred

@@

# Print table
table_fit1

# Save regression output into table
saveRDS(table_fit1, file = here::here("results", "tables", "table_fit1.rds"))
```

Now I will run a second simple bivariate analysis, looking at infant birth weight as a function of maternal age. This time, when plotting weight as a function of maternal age, I will use a scatterplot.

```{r}
# Model Age using recipes and workflows
# Create recipe
ML_recipe_age <- recipe(Weightgrams ~ Age, data = ML_linear)

@@

# Summarize model fit with tidiers
ML_summary_age <- summary(ML_fit_age) # Summary of fit
tidy(ML_fit_age, conf.int = TRUE) # Coefficients and confidence intervals
model_stats_age <- glance(ML_fit_age) # Extract model-level statistics

# This code is to write dynamic inline text in Quarto
# Extract R-squared value
r_squared_values$model2 <- model_stats_age$r.squared

# This code is for performance measures that aren't necessary in this study
# Produce predictions for RMSE and R-squared
ML_age_pred <- predict(ML_fit_age, new_data = ML_linear %>% select(-Weightgrams))
ML_age_pred

@@

Now I will run a slightly more complicated model that includes maternal skin color, age, nationality, and gestational status.

@@

# Summarize model fit with tidiers
ML_lm3_summary <- summary(ML_lm3) # Summary of fit
tidy(ML_lm3, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
model_stats3 <- glance(ML_lm3) # extract model-level statistics

# This code is to write dynamic inline text in Quarto
# Extract R-squared value
r_squared_values$model3 <- model_stats3$r.squared

@@

For all three models, the r-squared is small `r round(r_squared_values$model1, 3)`, `r round(r_squared_values$model2, 3)`, and `r round(r_squared_values$model3, 3)`, indicating that the model does not explain much of the variance in birth weight. This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors.

@@

## Sub-Sample Modeling

@@

```{r}
# Create a list to store R-squared values
r_squared_values2 <- list()
```

```{r}
# Model 4 sub-sample simple model
# Set reference level for Color
ML_linear$ModifiedNationality <- relevel(ML_linear$ModifiedNationality, ref = "Brazilian")

@@

# Summarize model fit with tidiers
ML_lm4_summary <- summary(ML_lm4) # Summary of fit
tidy(ML_lm4, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
model_stats4 <- glance(ML_lm4) # Extract model-level statistics

# Extract R-squared value
r_squared_values2$model4 <- model_stats4$r.squared

@@

# Save regression output into table
saveRDS(table_fit4, file = here::here("results", "tables", "table_fit4.rds"))
```

These results imply that Europeans had heavier infants than Brazilians. Since all women of color in this sample were Brazilian, there is a problem with multicollinearity. So I will perform some subsample modeling to understand this issue. We will create a `Brazilian` (yes, no) variable and then run a regression with this variable. Then, I will drop all non-Brazilians from the study and run the original regression with the covariates.

@@

str(ML_linear)
```

Now, I will run a regression with only this new variable.

```{r}
# Model 5 sub-sample simple model
# Set reference level for Brazilian
ML_linear$Brazilian <- relevel(ML_linear$Brazilian, ref = "Brazilian")

@@

# Summarize model fit with tidiers
ML_lm5_summary <- summary(ML_lm5) # Summary of fit
tidy(ML_lm5, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
model_stats5 <- glance(ML_lm5) # Extract model-level statistics

# Extract R-squared value
r_squared_values2$model5 <- model_stats5$r.squared

@@

# Save regression output into table
saveRDS(table_fit5, file = here::here("results", "tables", "table_fit5.rds"))
```

Here, non-Brazilians had babies on average weighing 100 grams more than Brazilians (p-value 0.001). So I will now run a regression of only Brazilian women using `Afro-Descent` and `Euro-descent` categories.

@@

# Model 6 sub-sample simple linear model
# Set reference level for Brazilian
ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")

@@

# Summarize model fit with tidiers
ML_lm6_summary <- summary(ML_lm6) # Summary of fit
tidy(ML_lm6, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
model_stats6 <- glance(ML_lm6) # Extract model-level statistics

# Extract R-squared value
r_squared_values2$model6 <- model_stats6$r.squared

@@

The $R^2$ for these models were `r round(r_squared_values2$model4, 3)`, `r round(r_squared_values2$model5, 3)`, and `r round(r_squared_values2$model6, 3)`.

## Multivariate analysis

Moving on from issues of collinearity, I will run a model where skin color is differentiated according to the original data, or with three categories (White, Black, and Mixed Race). I'll start with a simple model of infant birth weight as a function of maternal skin color. Then I will run a model with all the predictors.

```{r}
# Create a list to store R-squared values
r_squared_values3 <- list()
```

```{r}
# Model 7 simple linear model
# Set reference level for Color
ML_linear$Color <- relevel(ML_linear$Color, ref = "White")

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm7 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color, data = ML_linear)

# Summarize model fit with tidiers
ML_lm7_summary <- summary(ML_lm7) # Summary of fit
tidy(ML_lm7, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
model_stats7 <- glance(ML_lm7) # Extract model-level statistics

# Extract R-squared value
r_squared_values3$model7 <- model_stats7$r.squared

# Create a gtsummary table directly from the model
table_fit7 <- tbl_regression(ML_lm7, exponentiate = FALSE, intercept = TRUE) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))
table_fit7

# Save regression output into table
saveRDS(table_fit7, file = here::here("results", "tables", "table_fit7.rds"))

@@

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm8 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear)

# Summarize model fit with tidiers
ML_lm8_summary <- summary(ML_lm8) # Summary of fit
tidy(ML_lm8, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
model_stats8 <- glance(ML_lm8) # Extract model-level statistics

# Extract R-squared value
r_squared_values3$model8 <- model_stats8$r.squared

# Create a gtsummary table directly from the model
table_fit8 <- tbl_regression(ML_lm8, exponentiate = FALSE, intercept = TRUE) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))
table_fit8

# Save regression output into table
saveRDS(table_fit8, file = here::here("results", "tables", "table_fit8.rds"))
```

The $R^2$ for these models were `r round(r_squared_values3$model7, 3)` and `r round(r_squared_values3$model8, 3)`.

@@

Now I will still consider `Birthweight` as the outcome of interest, but I will use the dummy variable created in the `eda-v1.qmd` file that categorizes infant birth weight into normal birth weight (NBW $>2500$ g) or low birth weight (LBW $\leq2500$ g) and fit a logistic model, using the main predictor of interest, `ModifiedColor`.

First, I will split data into training and testing sets to later calculate ROC_AUC.

@@

Now I will run the first logistic model with `ModifiedColor` as the predictor.

```{r}
# Create recipe
# Logistic model 1: Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
log1_recipe <- recipe(BirthweightCategory ~ ModifiedColor, data = ML_linear) #%>%
  #step_dummy(all_nominal(), -all_outcomes()) %>%
  #step_normalize(all_predictors())

@@

# Print the plot
print(roc_plot1)

# Store metrics in a list for dynamic inline text
logistic_model_metrics1 <- list(
  pseudo_r_squared = mcfadden_r2_1,
  auc = roc_auc_result1$.estimate
)
```

Then, I will create a list of lists to store my performance metrics for dynamic inline text later.

```{r}
# Create a list to store performance metrics
all_logistic_models <- list()
all_logistic_models$model1 <- logistic_model_metrics1
```

With an ROC_AUC of `r round(all_logistic_models$model1$auc, 3)`, the model is performing worse than random guessing. The pseudo-$R^2$ is `r round(all_logistic_models$model1$pseudo_r_squared, 3)`.

Now I will run another logistic model with `Color` as the predictor.

```{r}
# Logistic model 2: simple logistic regression model with Color as predictor
# Check factor levels
levels(ML_linear$BirthweightCategory) # Reference is LBW

@@

# Extract model metrics using broom
model_metrics2 <- glance(log2_fitted)

@@

# Print the plot
print(roc_plot2)

# Store metrics in a list for dynamic inline text
logistic_model_metrics2 <- list(
  pseudo_r_squared = mcfadden_r2_1,
  auc = roc_auc_result1$.estimate
)
all_logistic_models$model2 <- logistic_model_metrics2 # add to list of lists
```

The ROC_AUC for the second simple logistic regression model is `r round(all_logistic_models$model2$auc, 3)`, or still worse than random guessing. The pseudo- $R^2$ is `r round(all_logistic_models$model2$auc, 3)`.

Now, I will run two more logistic models, one with `ModifiedColor` and the other covariates `Age`, `ModifiedStatus`, `ModifiedNationality`, and one with `Color` and these same covariates.

```{r}
# Logistic model 3: Multiple logistic regression model with ModifiedColor as the predictor
# Check factor levels
levels(ML_linear$BirthweightCategory) # Reference is LBW
levels(ML_linear$ModifiedColor) # Reference is Euro-descent
levels(ML_linear$ModifiedStatus) # Reference is multiparous
levels(ML_linear$ModifiedNationality) # Reference is European

@@

# Print the plot
print(roc_plot3)

# Store metrics in a list for dynamic inline text
logistic_model_metrics3 <- list(
  pseudo_r_squared = mcfadden_r2_1,
  auc = roc_auc_result1$.estimate
)
all_logistic_models$model3 <- logistic_model_metrics3 # add to list of lists
```

The ROC_AUC here is `r round(all_logistic_metrics$model3$auc, 3)`, still worse than random guessing. The pseudo-$R^2$ is `r round (all_logistic_metrics$model3$pseudo_r_squared, 3)`.

```{r}
# Logistic model 4: Multiple logistic regression model with Color as the predictor
# Check factor levels
levels(ML_linear$BirthweightCategory) # Reference is LBW
levels(ML_linear$Color) # Reference is White
levels(ML_linear$ModifiedStatus) # Reference is multiparous
levels(ML_linear$ModifiedNationality) # Reference is European

@@

# Print the plot
print(roc_plot4)

# Store metrics in a list for dynamic inline text
logistic_model_metrics4 <- list(
  pseudo_r_squared = mcfadden_r2_1,
  auc = roc_auc_result1$.estimate
)
all_logistic_models$model4 <- logistic_model_metrics4 # add to list of lists
```

In the final model, the ROC_AUC is `r round(all_logistic_metrics$model4$auc, 3)` still worse than random guessing. The pseudo-$R^2$ is `r round(all_logistic_metrics$model4$pseudo_r_squared, 3)`.

@@

# Save regression output into table
saveRDS(table5, file = here("results", "tables", "table5_final.rds"))
```

### 2024-08-16 code/eda-code/eda-v1.qmd
Revisions for submission 3
Clean up of code

Now let's make some tables about the data to get a better idea of the overall numbers.

Before we create a summary table, lets clean up our variables for easier analysis. First, we need to create dummy variables for categories. We will start with the `Color` variable, which is maternal skin color. We will also create dummy variables for the `Status` and `Nationality` variables for ease of interpretation. The original data categorized maternal racial categories as White (*branca*), the reference group here; mixed-race (*parda*); and Black (*preta*). I maintained this categorization. However, I also created a new dummy variable for skin color, combining Black and Mixed-Race patients into one category, `Afro-Descent` and all White patients into a `Euro-Descent` category.

For the `Status` variable, Table 1 will include all of the gravidity or parity categories from the original data. But I will also create a dummy variable for the linear analysis: one category for first pregnancy/birth and one for second or more pregnancies/births. With this, I am removing the distinction between parity and gravidity since it is unclear from the original data published in the clinical notes if the physicians were categorizing the women before the delivery or after. With this, I am assuming that both gravidity and parity refer to previous births, which risks including a history of stillbirths into a presumed history of live births. However, because these data are about women entering a hospital to give birth, rather than to receive prenatal care for a pregnancy, and this assumption can provide more statistical power for a linear regression, I assume this risk.

For first pregnancy/birth, I combine all "primi" statuses and "nulli" statuses (gravida or parous) into `nulliparous`. For second or more pregnancies/births, I combine all multiple statuses (gravida or parous) into `multiparous`. This combines "multigesta", "secundigesta", and "trigesta" (-gesta meaning -gravida in English) into `multiparous`, since they mean the same thing (two or more pregnancies). I also combine "secundipara" and "multipara" into the category `multiparous`, since this means two or more births. The words in Portuguese and English here are the same.

For the `Nationality` variable, we will combine all European nationalities into one category, "European", all Latin American nationalities into "LatinAmerican", and all Middle Eastern nationalities into "MiddleEastern". I am combining categories because for many of the `European` and `Latin American` categories, there are very few observations. Moreover, I am interested in immigration as a general status rather than the relationship between specific immigrant countries and health status in Brazil. "Brazilian" will remain the same.

Finally, for the summary table, in which I want to include infant birth weight stratified by maternal skin color and by fetal sex, I will create new dummy variables here for that purpose.

@@

# Verify new variables
str(ML_summary[, c("Weightblack", "Weightmixed", "Weightwhite", "Weightfemale", "Weightmale", "WeightgramsBlack", "WeightgramsMixed", "WeightgramsWhite", "WeightgramsFemale", "WeightgramsMale")])

# View the updated dataframe
print(head(ML_summary))

# See LBW and NBW with true denominator
LBW_percentage <- ML_summary %>%
# Remove NAs from both variables
filter(!is.na(ModifiedColor) & !is.na(BirthweightCategory)) %>%
# Group by ModifiedColor
group_by(ModifiedColor) %>%
# Calculate the percentage for each group
summarise(
  total_count = n(),
  lbw_count = sum(BirthweightCategory == "LBW"),
  lbw_percentage = (lbw_count / total_count) * 100
  )

# Print the result
print(LBW_percentage)

# Perform difference in means test (t-test) for birth weight category
# Modify code to create dataset for t-test
LBW_data <- ML_summary %>%
  filter(!is.na(ModifiedColor) & !is.na(BirthweightCategory)) %>%
  mutate(is_LBW = as.numeric(BirthweightCategory == "LBW"))  # 1 if LBW, 0 if not

# Print the first few rows to check the structure
print(head(LBW_data))

# Perform t-test
t_test_result <- t.test(is_LBW ~ ModifiedColor, data = LBW_data, subset = ModifiedColor %in% c("Euro-Descent", "Afro-Descent"))

# Print the result
print(t_test_result)
```

Let's also change the observations from Portuguese to English for the remaining observations to clean up things for the summary table.

@@

# Save new summary data with recoded factor variables
saveRDS(ML_summary, file = here("data", "processed-data", "ML_summary.rds"))

#Summarize missing data using dplyr
missing_data_summary <- ML_summary %>% dplyr::summarise_all(list(~sum(is.na(.))))
missing_data_summary
```

@@

# Save new summary data with recoded factor variables
saveRDS(ML_linear, file = here("data", "processed-data", "ML_linear.rds"))
```
# Notes

>>>>

Before we create a summary table, lets clean up our variables for easier analysis. First, we need to create dummy variables for categories. We will start with the `Color` variable, which is maternal skin color. We will also create dummy variables for the `Status` and `Nationality` variables for ease of interpretation. The original data categorized maternal racial categories as White (*branca*), the reference group here; mixed-race (*parda*); and Black (*preta*). I maintained this categorization. However, I also created a new dummy variable for skin color, combining Black and mixed-Race patients into one category, `Afro-Descent`, and all White patients into a `Euro-Descent` category.

For the `Status` variable, the Supplementary Table 1 will include all of the gravidity or parity categories from the original data. But I will also create a dummy variable for the summary table for both the final paper and the linear analysis: one category for first pregnancy/birth and one for second or more pregnancies/births. With this, I am removing the distinction between parity and gravidity since it is unclear from the original data if the physicians were categorizing the women before the delivery or after. With this, I am assuming that both gravidity and parity refer to previous births, which risks including a history of stillbirths into a presumed history of live births. However, because these data are about women entering a hospital to give birth, rather than to receive prenatal care for a pregnancy, and this assumption can provide more statistical power for a linear regression, I assume this risk.

For first pregnancy/birth, I combine all "primi" statuses and "nulli" statuses (gravida or parous) into `nulliparous`. For second or more pregnancies/births, I combine all multiple statuses (gravida or parous) into `multiparous`. This combines "multigesta", "secundigesta", and "trigesta" (-gesta in Portuguese meaning -gravida in English) into `multiparous`, since they mean the same thing (two or more pregnancies). I also combine "secundipara" and "multipara" into the category `multiparous`, since this means two or more births. The words in Portuguese and English here are the same.

For the `Nationality` variable, we will combine all European nationalities into one category, "European", all Latin American nationalities into "Latin American", and all Middle Eastern nationalities into "Middle Eastern". I am combining categories because for many of the `European` and `Latin American` categories, there are very few observations. Moreover, I am interested in immigration as a general status rather than the relationship between specific immigrant countries and health status in Brazil. "Brazilian" will remain the same.

Finally, for the summary table, in which I want to include infant birth weight stratified by maternal skin color and by fetal sex, I will create new dummy variables here for that purpose. For maternal skin color, these variables are `Weightblack`, `Weightmixed`, and `Weightwhite`. These are categorical variables (yes/no). From these, I create new continuous dummy variables for birth weight according to maternal skin color: `WeightgramsBlack`, `WeightgramsMixed`, and `WeightgramsWhite`. I do the same process for fetal sex: `Weightfemale`, `Weightmale`, `WeightgramsFemale`, and `WeightgramsMale`.

Using these new dummy variables, I create new dummy variables defining if the infants were normal birth weight (NBW $>2500$ g) or low birth weight (LBW $\leq2500$ g) for all infants and then stratified by maternal skin color. These are `Afro_NBW` and `Afro_LBW`; and `Euro_NBW` and `Euro_LBW`. I will use these to see how many infants were low birth weight or normal birth weight for each maternal skin category in the `introanalysis-v1.qmd` file.

@@

# Verify new variables
str(ML_summary[, c("Weightblack", "Weightmixed", "Weightwhite", "Weightfemale", "Weightmale", "WeightgramsBlack", "WeightgramsMixed", "WeightgramsWhite", "WeightgramsFemale", "WeightgramsMale", "Afro_LBW", "Afro_NBW", "Euro_LBW", "Euro_NBW")])

# View the updated dataframe
print(head(ML_summary))
```
Let's also change the observations from Portuguese to English for the remaining observations to clean up things for the summary table.

@@

#Summarize missing data using dplyr
missing_data_summary <- ML_summary %>% dplyr::summarise_all(list(~sum(is.na(.))))
missing_data_summary

# Save new summary data with recoded factor variables
saveRDS(ML_summary, file = here("data", "processed-data", "ML_summary.rds"))
```

@@

# Save new linear data with recoded factor variables
saveRDS(ML_linear, file = here("data", "processed-data", "ML_linear.rds"))
```
# Notes

### 2024-08-16 code/processing-code/processingfile-v1.qmd
Revisions for submission 3
Clean up of code

```{r}
#Summarize missing data using dplyr
missing_data_summary <- ML_4 %>% dplyr::summarise_all(list(~sum(is.na(.))))
missing_data_summary
```
Now, I will create a new categorical variable of Normal Birthweight (NBW) for all birth weights greater than 2500 grams and Low Birthweight (LBW) for all birth weights equal to or less than 2500 grams for a future logistic regression. This will come from the `Weightgrams` variable.

>>>>

```{r}
#Summarize missing data using dplyr
missing_data_summary <- ML_4 %>% dplyr::summarise_all(list(~sum(is.na(.))))
missing_data_summary
```
Now, I will create a new categorical variable of normal birth weight (NBW) for all birth weights $>2500$ g and low birth weight (LBW) for all birth weights $\leq2500$ g for a future logistic regression. This will come from the `Weightgrams` variable.
